### 2015-10-21 manuscript.Rnw
DJW edit
Mostly to intro and pretty minimal
 * Slightly more detail on growth of phylo, lack of synth. at present
 * Increased emph. on R/rotl as reproducable research
 * Mention published trees (not just synth) in inro

Other minor changes
 * Added comment about accepting features/bugs in availability
 * Short blurb about tests/docs in technical details

%%% syst bio options
\linespread{1.66}
% All text should be double-spaced
% with occasional exceptions for tables.
\raggedright
\setlength{\parindent}{0.5in}

@@

% Our sections are not numbered and our papers do not have
% Tables of Contents. We don't
% present a list of figures or list of tables, either.

% Any common font is fine.
% (A common sans-serif font should be used on figures, but figures should be
% separate from the LaTeX document.)

@@

% ----------------------------------------------------------------------------- %

@@

Generating phylogenies has been greatly facilitated by advances in sequencing
and computing technologies. In parallel, new phylogenetic comparative methods
have been developed and make use of these phylogenies to explore fundamental
questions about the origin of biodiversity including the evolution of
morphological and ecological traits, the spatio-temporal variation in speciation
rates, or both \citep{OMeara2012,Pennell2013}. The development of these methods
has contributed to the need for a greater availability of high-quality
phylogenetic information across life. However, phylogenetic information is
scattered, often only available as image files within publications, and the lack
of standardization to store and represent phylogenetic data makes it difficult
for researchers to access, synthesize, and integrate this information into their
own research (\citealt{Stoltzfus2012}, but see \citealt{Cranston2014} for
suggestions of best practices).

@@

The Open Tree of Life project provides a web interface to explore, query and
download parts of the synthetic Open Tree
(\url{http://tree.opentreeoflife.org}). Additionally, the project provides
Application Programming Interfaces (APIs) to access programmatically the
different services and underlying data. Here, we present \texttt{rotl} a package
to interact with OTL's APIs to search for and import phylogenetic trees and
taxonomic information directly in R. Having access to this data in R allows
researchers to take advantage of the multiple packages that are available to
visualize, analyze and manipulate phylogenetic and comparative data (e.g., ape
\citealt{Paradis2004}, phytools \citealt{Revell2012}, geiger
\citealt{Pennell2014-geiger}, ggtree \citealt{Yu-ggtree}, phylobase
\citealt{phylobase-0.8.0}, RNeXML \citealt{Boettiger2015}, see
\url{https://cran.r-project.org/web/views/Phylogenetics.html} for a
comprehensive list). Furthermore, \texttt{rotl} joins a growing list of packages
that aim at querying and accessing data from the web directly in R for analysis.

@@

@@

\texttt{rotl} is free, open source, and released under a Simplified
BSD license. Stable versions are available from the CRAN repository
\url{https://cran.r-project.org/package=rotl}, and development
versions are available from GitHub
\url{https://github.com/ropensci/rotl}. The source for this manuscript is
available at \url{https://github.com/fmichonneau/rotl-ms}.

>>>>

%%% syst bio options \linespread{1.66}
% All text should be double-spaced with occasional exceptions for tables.
\raggedright \setlength{\parindent}{0.5in}

@@

% Our sections are not numbered and our papers do not have Tables of Contents.
% We don't present a list of figures or list of tables, either.

% Any common font is fine.  (A common sans-serif font should be used on figures,
% but figures should be separate from the LaTeX document.)

@@

% -----------------------------------------------------------------------------
% %

@@

Advances in sequencing and computing technologies have lead to a revolution in
systematic biology. The ability to routinely generate molecular datasets from
any extant organism has allowed researchers to resolve long-running taxonomic
disputes and estimate phylogenies for previously understudied groups. In
parallel, the ease which which phylogenies can be estimated has lead to the
development of new phylogenetic comparative methods. These methods allow
researchers to explore fundamental questions about the origin of biodiversity
including the evolution of morphological and ecological traits, the
spatio-temporal variation in speciation rates, or both
\citep{OMeara2012,Pennell2013}.

Ideally, the ever increasing number of published phylogenies would contribute to
a synthesis of phylogenetic knowledge -- leading to a better understanding of
the history of life and providing high-quality phylogenetic information for use
in comparative methods. However, in practice synthesizing phylogenetic data
remains difficult. Phylogenies are scattered, often only available as image
files within publications, and the lack of standardization to store and
represent phylogenetic data makes it difficult for researchers to access,
synthesize, and integrate this information into their own research
(\citealt{Stoltzfus2012}, but see \citealt{Cranston2014} for suggestions of best
practices).

@@

The Open Tree of Life project provides a web interface to explore, query and
download parts of the synthetic Open Tree (\url{http://tree.opentreeoflife.org})
and the published trees that contribute to it. Additionally, the project
provides Application Programming Interfaces (APIs)  to access programmatically
the different services and underlying data.  Here, we present \texttt{rotl} a
package to interact with OTL's APIs to search for and import phylogenetic trees
and taxonomic information directly in R. Having access to this data in R allows
researchers to take advantage of the multiple packages that are available to
visualize, analyze and manipulate phylogenetic and comparative data (e.g., ape
\citealt{Paradis2004}, phytools \citeateu{Revell2012}, geiger
\citealt{Pennell2014-geiger}, ggtree \citealt{Yu-ggtree}, phylobase
\citealt{phylobase-0.8.0}, RNeXML \citealt{Boettiger2015}, see
\url{https://cran.r-project.org/web/views/Phylogenetics.html} for a
comprehensive list). There is also a growing list of packages that allow users
to query and acces data from the web directly in R \hl{rFISHBASE
10.1111/j.1095-8649.2012.03464.x, paleoDB 10.1111/ecog.01154, rAvis
10.1371/journal.pone.0091650 \dots}. By providing direct access to high quality
phylogenetic data in R, \texttt{rotl} fills a key gap typical comparative
analysis workflows. Moreover, by allowing the gathering, processing and analysis
of data to be performed programmatically, \texttt{rotl} facilitates reproducible
research practices.

@@

The package is well-documented, and includes two package
vignettes (documents that demonstrate the use of the package and contain executable 
R code) as well as per-function documentation. There is also an extensive
test-suite that covers both the internal functions that \textt{rotl} uses to
connect to OTL and public functions users apply to acess and process data.

@@

\texttt{rotl} is free, open source, and released under a Simplified
BSD license. Stable versions are available from the CRAN repository
\url{https://cran.r-project.org/package=rotl}, and development
versions are available from GitHub
\url{https://github.com/ropensci/rotl}. The package is under active development,
and authors welcome bug reports of feature requests via the github repository.
The source for this manuscript is available at \url{https://github.com/fmichonneau/rotl-ms}.
